---
title: "Getting Started with Relational Data"
author: "Jessi Kehe"
output: html_document
---

### Setup details

* You will need the packages **tidyverse**, **lubridate**, **nasaweather** for this lecture. 

* Be sure to read R4DS Chapter 13 for information on relational data

* We are going to consider a few simple examples to help you get started with relational data.  Practice on more complicated examples will come during the lectures and as the semester progresses.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
library(lubridate)
install.packages("nasaweather")
library(nasaweather)
```

### Keys

A **primary key** for a data frame are the variable or variables that uniquely identify an observation.

Let's try to find the primary key in the `nasaweather::atmos` data.  We will begin by running `?nasaweather::atmos` for details on the data, which may provide some hint at a primary key.

```{r}
?nasaweather::atmos

atmos <- nasaweather::atmos
class(atmos)
dim(atmos)
names(atmos)

atmos %>% 
  count(lat, long, year, month) %>% 
  filter(n > 1)

atmos %>% 
  count(lat,long,year,month) %>% 
  filter(n > 1)
```


### Mutating joins

We were introduced to joins in week 3 in our introduction to dplyr (see `03-intro-dplyr.Rmd`), but we will consider another example here.


Consider the following two data frame. 

```{r drinks-discounts}
drinks <- tibble(
  Drink = c("Milk Tea", "Espresso", "Latte", "Espresso"),
  Cafe = c("Happy Tea", "Daily Grind", "Daily Grind", "Bev's Beverages"),
  Price = c(4, 2, 3, 2)
)
drinks

discounts <- tibble(
  Coupon = c("25%", "50%", "5%"),
  Location = c("Happy Tea", "Daily Grind", "The Roast")
)
discounts
```


```{r mutating-joins}
drinks %>% 
  inner_join(discounts, by = c("Cafe" = "Location"))

drinks %>% 
  full_join(discounts, by = c("Cafe" = "Location"))

drinks %>% 
  full_join(discounts, by = c("Cafe" = "Location"), keep=TRUE)

drinks %>% 
  left_join(discounts, by = c("Cafe" = "Location"))

drinks %>% 
 right_join(discounts, by = c("Cafe" = "Location"))
```


### Filtering joins

Filtering joins also look to match observations between the two data frame, but focus on the observations instead of the variables. 

There are two filtering joins:  `semi_join()` and `anti_join()`.  

```{r filtering-joins}
# semi_join() returns all observations of drinks with a match in discounts
drinks %>% 
  semi_join(discounts, by = c("Cafe" = "Location")) 

discounts %>% 
  semi_join(drinks, by = c("Location"="Cafe"))



# anti_join() returns all observations of drinks that do *not* have a match in discounts
drinks %>% 
  anti_join(discounts, by = c("Cafe" = "Location"))

discounts %>% 
  anti_join(drinks, by = c("Location"="Cafe"))
```

