---
title: "Getting Started with tidyr"
author: "Jessi Kehe"
output: html_document
---

### Setup details

* You will need the packages **tidyverse**, **lubridate**, and **XML** for this lecture. 

* The following data files will be used and are assumed to be located two steps back from the working directory in a folder called `data/`.  Be sure to adjust the code if you have the data files in a different location.  
`"../../data/DEN-2018.html"`  
`"../../data/JFK-2018.html"`  
`"../../data/LAX-2018.html"`  
`"../../data/MSP-2018.html"`  
`"../../data/ORD-2018.html"`  
`"../../data/SFO-2018.html"` 

* Be sure to read R4DS Chapter 12 for information on **tidyr** 

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
library(lubridate)
library(XML)
```


# tidyr 

Tidy data satisfy the following criteria

  1. Each variable has its own column  
  2. Each observation has its own row
  3. Each value has its own cell
  

## Airport Data

Let's consider our airport timing data from last week.

```{r stuff-for-all-airports}
# Code to read in the data files
read_airport <- function(x)
  # Input:  x = HTML file
  # Output: nicely formatted tibble
{
  airport_column_names <- c("airport",
                           "terminal",
                           "date",
                           "hour",
                           "us_avg_wait",
                           "us_max_wait",
                           "non_us_avg_wait",
                           "non_us_max_wait",
                           "all_avg_wait",
                           "all_max_wait",
                           "all_n_0_15",
                           "all_n_16_30",
                           "all_n_31_45",
                           "all_n_46_60",
                           "all_n_61_90",
                           "all_n_91_120",
                           "all_n_120_plus",
                           "all_excluded",
                           "all_total",
                           "all_flights",
                           "all_booths")

  airport <- readHTMLTable(x)$`NULL` %>%
    slice(-(1:3))

  names(airport) <- airport_column_names

  airport <- airport %>%
    as_tibble() %>%
    mutate_at(.vars = vars(ends_with("wait"),starts_with("all")),
              .funs = as.numeric) %>%
    mutate_if(is.factor,as.character) %>%
    mutate(date = mdy(date))
  
  return(airport)
}
```

### Read the Data

```{r read-data}
den <- read_airport("C:/stat_240/data/DEN-2018.html")
jfk <- read_airport("C:/stat_240/data/JFK-2018.html")
lax <- read_airport("C:/stat_240/data/LAX-2018.html")
msp <- read_airport("C:/stat_240/data/MSP-2018.html")
ord <- read_airport("C:/stat_240/data/ORD-2018.html")
sfo <- read_airport("C:/stat_240/data/SFO-2018.html")
```

```{r combine-data}
awt <- bind_rows(den,jfk,lax,msp, ord,sfo)
rm(den,jfk,lax,msp,ord,sfo)

str(awt)
```


### pivot_longer()

- Each observation in the data set is a single airport/terminal/date/hour.

- However, some of the columns contain data.
    - Namely, the `all_n_*` columns contain information in the names about the number of passengers who wait for given time intervals.
    
- An alternative way to represent the data would be to have all of these counts be in a single column.

- Then, an observation would entail a airport/terminal/date/hour/wait-interval combination.

```{r pivot-longer}
names(awt)

awt %>%
  pivot_longer(cols=starts_with("all_n_"),names_to="wait_interval",names_prefix="all_n_",values_to="passenger_count") %>%
  select(airport,terminal,date,hour,wait_interval,passenger_count,everything())

```










