---
title: "Madison Crime Data"
author: "Jessi Kehe"
output: html_document
---
This R Markdown document was adapted from documents created by Professor Bret Larget.

### Setup details

* You will need the packages `tidyverse`, `lubridate`, and `stringr` for these lectures.  Make sure to install them if you haven't already.  

* This assumes you have the R script `viridis.R` two steps back from the working directory (`"../../scripts/viridis.R"`).  Be sure to adjust the code if you have this script in a different location.

* The following data files will be used and are assumed to be located two steps back from the working directory in a folder called `data/`.  Be sure to adjust the code if you have the data files in a different location.  
`"../../data/Police_Incident_Reports_2020-09-28.csv"`  


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
library(lubridate)
## Note: this code assumes viridis.R is two steps up the file directory tree in a `scripts` folder
source("C:/stat_240/scripts/viridis.R")
```


# Lecture:  Madison Crime Data Exploration


## Madison Police Incident Reports

- The data set is a summary of police incident reports in the city of Madison, Wisconsin.

- The source of the data is [http://data-cityofmadison.opendata.arcgis.com/datasets/police-incident-reports](http://data-cityofmadison.opendata.arcgis.com/datasets/police-incident-reports)

- The data website notes, "Incidents listed are selected by the Officer In Charge of each shift that may have significant public interest. Incidents listed are not inclusive of all incidents."  
  - This means we do not have a census of incidents nor a random sample so we should be cautious about drawing conclusions from these data

- The data may be downloaded as a CSV file.

- The web page describes these variables:
    - Address
    - Arrested
    - CaseNumber
    - DateModified
    - Details
    - IncidentDate
    - IncidentID
    - IncidentType
    - ReleasedBy
    - Suspect
    - Victim



```{r read-data}
police_orig <- read_csv("C:/stat_240/data/Police_Incident_Reports_2020-09-28.csv")

head(police_orig)
str(police_orig)


# Need to format the date variables
police_orig <- police_orig %>%
  mutate(IncidentDate = as_datetime(IncidentDate)) %>% # Format as_datetime
  mutate(DateModified = as_datetime(DateModified)) 

head(police_orig) # POSIXct = date-time 
str(police_orig)
```

### Initial Explorations


> Check if the `IncidentID` uniquely identifies an observation

```{r distinct-id}
police_orig %>%
  nrow()

police_orig %>%
  select(IncidentID) %>%
  distinct() %>%
  nrow()
```

- Yes, all incident IDs are distinct



>  What are the different `IncidentType` values?  That is, what types of incidents are reported?

```{r IncidentType}
police_orig %>%
  count(IncidentType) %>%
  arrange(desc(n)) %>%
  print(10)

police_orig %>%
  count(IncidentType) %>%
  slice_max(order_by=n,n=10)

police_orig %>%
  count(IncidentType) %>%
  slice_min(order_by=n,n=10)
```

- 99 types
- most frequent is Robbery, 2563 times



> How many case-numbers match the pattern 4 digits - 1 or more digits?  

```{r count-case-number-pattern}
police_orig %>%
  mutate(case_match = 
           # str_detect(CaseNumber,"^[0-9]+-[0-9]?$")) %>%  # Matches one+ digits - zero or one digits
           str_detect(CaseNumber,"^[0-9]+-[0-9]{0,1}$")) %>%
  filter(case_match) 


police_orig %>%
  mutate(case_match = 
           # str_detect(CaseNumber,"^[0-9]+-[0-9]+$")) %>%  # Matches one+ digits - one+ digits
           # str_detect(CaseNumber,"^[0-9]{4}-[0-9]+$")) %>%  # Matches four digits - one+ digits
           str_detect(CaseNumber,"^[0-9]{4}-[0-9]{5,6}$")) %>% # Matches four digits - five-six digits
           # str_detect(CaseNumber,"^[0-9]{4}-[0-9]{6}$")) %>% # Matches four digits - six digits
  filter(!case_match) %>%
  select(IncidentID,CaseNumber,IncidentDate)
```

- It is apparent that a valid case number is a four digit number matching the year, followed by a dash, and then a number of digits, usually 5 or 6

- Patterns that do not match this can mismatch in a number of different ways.

- Some should be corrected


> What is the range of dates?

```{r date}
# Look at date range
police_orig %>%
  summarize(min = min(IncidentDate),
            max = max(IncidentDate))

police_orig %>%
  slice_max(n=10, order_by=IncidentDate) 

police_orig %>%
  slice_min(n=10,order_by=IncidentDate)
```

- There appears to be some problems with some of the dates


> How does the case number relate to the date of the incident? Are there any errors?

```{r explore-IncidentID-order}
date_temp <- police_orig$IncidentDate[1]
date_temp
date_temp - now()

## See if IncidentID is chronological
ggplot(police_orig, aes(x=IncidentDate-now(),y=as.numeric(IncidentID))) +
  geom_point()
```

- As IncidentDate increases, the IncidentID generally increases  
- We observe several points that do not follow this pattern










# Lecture:  Incident Dates and Case Numbers

- We saw that one date is in the future and a few are far in the past.

- Let's flag cases where `IncidentDate` is after the present or the year does not match the first four characters of the `CaseNumber`.

```{r clean-dates}
## There are some clear outliers where the IncidentDate appears to be wrong.

bad_date <- police_orig %>%
  filter(IncidentDate > now() | # dates in the future
           str_sub(CaseNumber,1,4) != year(IncidentDate)) %>% # first part of case number does not match incident date
  select(IncidentID,CaseNumber,IncidentDate,DateModified)
bad_date


## Change all CaseNumbers which begin 0X- to 200X-
## Change CaseNumbers which begin with two or three digits and then a dash
##   to match the year of the date before the dash
## Turn "none"'s into NAs
## Turn CaseNumbers that have only zeros after the - into NAs
police <- police_orig %>%
  mutate(CaseNumber = case_when(str_detect(CaseNumber,"^0[0-9]-") ~ str_c("20",CaseNumber), 
    str_detect(CaseNumber,"^[0-9]{2}-") ~ str_c(year(IncidentDate),str_sub(CaseNumber,3,-1)),
    str_detect(CaseNumber,"^[0-9]{3}-") ~ str_c(year(IncidentDate),str_sub(CaseNumber,4,-1)),
    str_detect(str_to_lower(CaseNumber),"none") ~ as.character(NA), 
    TRUE ~ CaseNumber)) %>%
  mutate(CaseNumber = case_when(str_detect(CaseNumber,"-0+$") ~ as.character(NA),
    TRUE ~ CaseNumber)) 

police %>%
  filter(IncidentDate > now() |
           str_sub(CaseNumber,1,4) != year(IncidentDate)) %>%
  select(IncidentID,CaseNumber,IncidentDate,DateModified)
## Still issues remaining; would want to ask if year in IncidentDate is more
## reliable than the first four digits of the CaseNumber
## If so, we could fix many of these issues

police %>%
  slice_max(n=1, IncidentDate) #This IncidentDate likely could get corrected to 2017
```










#  Lecture:  Incident frequency by day of the week

> Do the number of incidents vary by day of the week?  by month?

```{r day}
week <- police %>%
  select(IncidentID, IncidentType, IncidentDate) %>%
  mutate(wday = wday(IncidentDate, label = TRUE, abbr = TRUE),
         month = month(IncidentDate, label = TRUE, abbr = TRUE))

week %>%
  count(wday)%>%
  ggplot(aes(wday,y=n)) +
  geom_col()

week %>%
  count(month)%>%
  ggplot(aes(month,y=n)) +
  geom_col()
```

#### Does it vary for traffic-related incidents?

```{r}
week %>%
  select(IncidentType) %>%
  filter(str_detect(IncidentType,"(?i)traffic|drive|accident")) %>%
  distinct()

week %>%
  filter(str_detect(IncidentType,"(?i)traffic|drive|accident")) %>%
  count(wday)%>%
  ggplot(aes(wday,y=n)) +
  geom_col()

week %>%
  filter(str_detect(IncidentType,"(?i)traffic|drive|accident")) %>%
  count(month)%>%
  ggplot(aes(month,y=n)) +
  geom_col()
```

* You can explore other incident types in a similar manner.










# Lecture:  Incident Types and Times of Day


> How do the number of incidents change throughout the day?

We can define different windows of time to analyze the number of incidents:

“overnight” [1am to 5am)  
“early morning” [5am to 9am)   
“morning” [9am to 1pm)   
“afternoon” [1pm to 5pm)   
“evening” [5pm to 9pm)   
“late evening” [9pm to 1am)  

```{r incident-time}
it <- police %>%
  select(IncidentID, IncidentType, IncidentDate) %>%
  mutate(hour = hour(IncidentDate),
         time = case_when(
           hour < 1 ~ "late evening",
           hour < 5 ~ "overnight",
           hour < 9 ~ "early morning",
           hour < 13 ~ "morning",
           hour < 17 ~ "afternoon",
           hour < 21 ~ "evening",
           TRUE ~ "late evening")) 

## Which time has the most incidents
it %>%
  count(time) %>%
  arrange(desc(n))

ggplot(it, aes(x=time)) +
  geom_bar() +
  xlab("Incident Type") +
  ylab("Count") +
  ggtitle("Madison Crime Data")

## The barplot is helpful...but the order is a bit difficult to follow
### how the number of incident change across the day
## Let's reorder the time variable

it2 <- it %>%
  mutate(time_order = case_when(
           hour < 1 ~ 6,
           hour < 5 ~ 1,
           hour < 9 ~ 2,
           hour < 13 ~ 3,
           hour < 17 ~ 4,
           hour < 21 ~ 5,
           TRUE ~ 6), 
         time = reorder(time, time_order)) 
str(it2)


ggplot(it2, aes(x=time)) +
  geom_bar() +
  xlab("Incident Type") +
  ylab("Count") +
  ggtitle("Madison Crime Data")
```


> Do the *types* of incidents change throughout the day?

We will put the incident types into some broader categories, and then plot against hour.

```{r incident-types}
it %>%
  group_by(IncidentType) %>%
  summarize(n = n()) %>%
  arrange(desc(n))

## other types of robbery?
it %>%
  select(IncidentType) %>%
  filter(str_detect(IncidentType, "(?i)robbery")) %>%
  distinct()

## other types of traffic issues?
it %>%
  select(IncidentType) %>%
  filter(str_detect(IncidentType,"(?i)traffic|drive|accident")) %>%
  distinct()

## other types of burglaries?
it %>%
  select(IncidentType) %>%
  filter(str_detect(IncidentType,"(?i)burglar")) %>%
  distinct()

## other types of animals?
it %>%
  select(IncidentType) %>%
  filter(str_detect(IncidentType,"(?i)animal")) %>%
  distinct()

it3 <- it2 %>%
  mutate(incident_group = 
           case_when(
             str_detect(IncidentType,"(?i)Robbery") ~ "robbery",
             str_detect(IncidentType,"(?i)Weapons") ~ "weapons",
             str_detect(IncidentType,"(?i)Battery") ~ "battery",
             str_detect(IncidentType,"(?i)Theft") ~ "theft",
             str_detect(IncidentType,"(?i)traffic|drive|accident") ~ "traffic",
             str_detect(IncidentType,"(?i)burglary") ~ "burglary",
             TRUE ~ "other"
           ))

it3 %>%
  group_by(incident_group, hour) %>%
  summarize(n = n()) %>%
  ggplot(aes(x=hour,y=n,color=incident_group)) +
  geom_line(size=2) +
  xlab("Hour") +
  ylab("Count") +
  ggtitle("Madison Crime Data")


it3 %>%
  filter(incident_group == "other") %>%
  count(IncidentType) %>%
  arrange(desc(n))


it3 %>%
  group_by(incident_group, hour) %>%
  summarize(n = n()) %>%
  filter(incident_group != "other") %>%
  ggplot(aes(x=hour,y=n,color=incident_group)) +
  geom_line(size=2) +
  xlab("Hour") +
  ylab("Count") +
  ggtitle("Madison Crime Data")
```










# Lecture: Investigating the variable `Arrested`

## Examine a Sample of Arrested Entries


Suppose we want to develop a function that attempts to determine the race of the person arrested.

Let's look at a random sample of the suspects.

```{r sample}
## set.seed to be able to replicate the random sample
set.seed(1)
police_1 <- police_orig %>%
  filter(!is.na(Arrested)) %>%
  sample_n(size=100) %>%
  select(IncidentID,Arrested) %>%
  arrange(as.numeric(IncidentID))
print(police_1, n=100)
```


## Start building a function to return race

The function below is only to get started - more coding is needed to get it to work as expected.  
  - Eventually, the last line should return `as.character(NA)` instead of "X"  
  - To make the function more accurate, add code or modify code to handle new patterns that the existing code does not detect or to correct errors in the existing code.  

```{r race-function}
## attempt to extract race from an unformatted string
## in the Madison Crime Incident data
## x is the input string
extract_race <- function(x)
{
  x_lower <- str_to_lower(x)
  df <- tibble(x) %>%
    mutate(race = case_when(
      is.na(x) ~ as.character(NA),
      str_detect(str_to_lower(x),"^n/a$") ~ as.character(NA),
      str_detect(x_lower,"^unknown$") ~ as.character(NA),
      str_detect(x_lower,"white,") ~ "white",
      str_detect(x_lower,"[mf]/w") ~ "white",
      str_detect(x_lower,"w/[mf]") ~ "white",
      str_detect(x_lower,"black,") ~ "black",
      str_detect(x_lower,"[mf]/b") ~ "black",
      str_detect(x_lower,"b/[mf]") ~ "black",
      str_detect(x_lower,"african") ~ "black",
      str_detect(x_lower,"asian,") ~ "asian",
      str_detect(x_lower,"hispanic,") ~ "hispanic",
      str_detect(x_lower,"[mf]/h") ~ "hispanic",
      str_detect(x_lower,"h/[mf]") ~ "hispanic",
      str_detect(x_lower,"latin[oax]") ~ "latinx",
      TRUE ~ "X"
    ))
  return( pull(df,race) )
}

police_1_test <- police %>%
  mutate(suspect_race = extract_race(Arrested))

check_id <- 8683 #8977 #8683 #8170
police_1_test %>%
  filter(IncidentID == check_id) %>%
  pull(Arrested)

police_1_test %>%
  select(IncidentID, suspect_race) %>%
  filter(IncidentID == check_id)

police_1_test %>%
  select(suspect_race) %>%
  table()
```


































