---
title: "Getting Started with stringr"
author: "Jessi Kehe"
output: html_document
editor_options: 
  chunk_output_type: inline
---

This R Markdown document was adapted from documents created by Professor Bret Larget.


### Setup details

* You will need the packages **tidyverse** and **stringr** for this lecture. (**stringr** should load with **tidyverse**, but if not, run `library(stringr)` separately)

* Be sure to read R4DS Chapter 14 for information on **stringr** 

  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
```


# Lecture:  Strings and Regular Expressions

## Overview

This lecture will give examples of working with strings and regular expressions.

### Data

```{r color-data}
df_colors = tibble(
  index = seq_along(colors()), #length of seq based on length of argument
  colors = colors()) #colors() are built-in color names 

df_colors
```

## Problems

> Find the shortest and longest color names

- use `str_length()` to find the length of each string  
- then use `slice_max()` or `slice_min()` to identify the extremes  

```{r short-long}
df1 <- df_colors %>%
  mutate(length = str_length(colors))
head(df1)

df1long <- df1 %>%
  slice_max(n=1, order_by=length)
df1long

df1short <- df1 %>%
  slice_min(n=1, order_by=length)
df1short

```

- The longest color name is `r df1long$colors` with a length of `r df1long$length` characters.  
- The shortest color names are `r df1short$colors[1]` and `r df1short$colors[2]`, each with a length of `r df1short$length[1]`.



> Make a bar plot of the string lengths

- Note use of `scale_x_continuous()` to change the default breaks and labels in the plot.

```{r length-plot}
ggplot(df1, aes(x=length)) +
  geom_bar(fill=df1long$colors,color="black") +
  xlab("Word Length") +
  scale_x_continuous(labels=3:20,breaks=3:20) +
  ggtitle("R Colors Name Length Distribution")
```



> Find the color names of length 8. Report the number, and the first and last color name (alphabetically).

```{r length-8}
df_colors %>%
  filter(str_length(colors) == 8) %>%
  summarize(n = n(),
            first = min(colors),
            last = max(colors))

#Look at all colors with string length 8 in order by color name
df_colors %>%
  filter(str_length(colors) == 8) %>%
  arrange(colors)
```

> How many color names begin and end with the same letter? Report the number and the first and last alphabetically.

- `str_sub(string, start, end)` returns the sub string from `start` to `end`
    - Use negatives to count from the end.
    - Can also be used to assign values by position
    
```{r same-first-last}
df2 <- df_colors %>%
  mutate(first = str_sub(colors,1,1),
         last = str_sub(colors,-1,-1)) %>%
  filter(first == last)
df2

df2 %>%
  summarize(n=n(), first = min(colors), last = max(colors))
```



##### Here is a fancy regular expression ("regexp") to find colors that begin and end with the same letter

```{r same-first-last-fancy}
df2b <- df_colors %>%
  filter(str_detect(colors, "(^[a-zA-Z]).*\\1$"))

#Read carefully chapter 14.3
#\\1 - last character

df2b
```


> Find all color names that begin with a letter from "aeiou" and end with a letter from "aeiouy". Modify so that the first letter is capitalized and the last two letters are each replaced by X.

- Regular expression special symbols:
    - ^ is the beginning of the string
    - $ is the end of the string
    - . is any character (except a newline)
    - * means repeat 0 or more times
    - [...] matches any character from between the [ and ]
- `str_c()` combines string arrays, element by element
- `str_detect()` returns TRUE or FALSE that it detects the pattern in the string
- `str_to_upper()` changes a letter to upper case
- `str_sub()` may also be used for assignment
    
```{r vowel}
df3 <- df_colors %>%
  filter(str_detect(colors, "^[aeiou].*[aeiouy]$"))

df3

df3 <- df3 %>%
  mutate(new_colors = str_c(str_to_upper(str_sub(colors,1,1)),
                            str_sub(colors,2,-3), #start and end points are inclusive
                            "XX"))

df3

```

> Find all colors names which begin with a vowel, end with a consonant, have no digits in the name, and have between 4 and 10 total characters

- Regular expression `\d` stands for a digit.
- In an R string, need to double the `\` in order to represent the literal `\`.
- A ^ right after [ negates the matches between the [ and ]
- {n} repeats the previous pattern n times
- {m,n} repeats the previous pattern from m to n times

```{r vowel-consonant}
df4 <- df_colors %>%
  filter(!str_detect(colors,"\\d")) %>% # no digit
  filter(str_detect(colors, "^[aeiou].{2,8}[^aeiouy]$")) %>%
  mutate(length = str_length(colors)) # added to check the length
df4
```

> Find all color names that repeat a letter twice in a row.

- Surrounding a pattern with () changes the order of preference and forms a group.
- Whichever pattern matches a group may be referred to by `\x` where `x` is the number of the group
    - For example, `([a-e])\1` will match double letters from a through e.
    - But in a string in R, we need to precede `\` by another `\` to indicate we mean the literal `\`. So, `"([a-e])\\1"`
    
```{r repeat}
df5 <- df_colors %>%
  filter(str_detect(colors, "([a-zA-Z])\\1")) #`\\1` is called a backreference
df5
```

> Find all color names which contain the substrings "red", "blue", or "green"

- separate alternative patterns with a `|`

```{r rgb}
df6 <-df_colors %>%
  filter(str_detect(colors, "red|blue|green"))
df6
```








