---
title: "Text Analysis with stringr"
author: "Jessi Kehe"
output: html_document
editor_options: 
  chunk_output_type: inline
---

### Setup details

* You will need the packages **tidyverse** and **stringr** for this lecture. (**stringr** should load with **tidyverse**, but if not, run `library(stringr)` separately)

* We will also be using the packages **gutenbergr** in order to obtain some data so be sure to install it before moving forward.

  
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
library(gutenbergr)
library(tidytext)

```

## Overview

We are going to explore a text from **gutenbergr**, which is an R package that provides an interface for 
downloading public domain texts in the Project Gutenberg collection.  


## Data: gutenbergr

```{r data}
# Look at the structure of the gutenberg meta data
str(gutenberg_metadata)
```

```{r gutenbergr-titles}
gutenberg_metadata %>%
  select(title)
```


With over 50,000 texts it would be hard to look through to find a text we want.  
Fortunately we have `dplyr` and `stringr` to help us!

Let's find a title of a text...how about Huck Finn?  Now what is the full title?
I can't remember.  Instead, let's do a search for `huck` and see if we can find a match.

```{r huck}
gutenberg_metadata %>%
  filter(str_detect(title, "huck"))
```

Weird, this classic seems to not be in here.  Maybe some letters should be capitalized?

```{r huck-2}
gutenberg_metadata %>%
  select(gutenberg_id, title) %>%
  filter(str_detect(title, regex("huck", ignore_case=TRUE)))

gutenberg_metadata %>%
  select(gutenberg_id, title) %>%
  filter(str_detect(title, "(?i)huck" )) # (?i) makes it match case insensitive

#There it is!  Let's grab the work using the gutenberg_id = 76

huck <- gutenberg_download(76)
```


## Little Women

```{r get-lw}
# Louisa May Alcott authored many texts!
gutenberg_metadata %>%
  select(gutenberg_id, title, author) %>%
  filter(str_detect(author, "(?i)alcott"))

# We will look at little women
gutenberg_metadata %>%
  filter(str_detect(title, "^(?i)little women$"))
# gutenberg_id = 514...other option does not seem to have a text available

lw_orig <- gutenberg_download(514)
```


> Make a data frame with a chapter index, chapter titles, and the text

```{r chapters}
lw_orig # not very clean

lw_orig %>%
  slice(1:69) #first 69 rows are header material (like chapter names)

## Get chapter names
chapter_names <- lw_orig %>%
  select(text) %>%
  slice(c(16:38,43:60,61:66)) %>%
  mutate(text = str_trim(text)) %>%
  pull()
chapter_names

## Chapter 1 begins in row 70 and then the chapter content is spread across multiple rows.
lw_orig %>%  
  slice(70:80)

## Collapse the text into one string
lw <- lw_orig %>% 
  slice(-c(1:69)) %>% # remove header content
  select(text) %>%
  filter(!str_detect(text,"^$")) %>% #remove the empty strings
  summarize(full = str_c(text, collapse = " ")) 

length(lw$full)
str_length(lw$full)

## Split the text into a string for each chapter
little_women <- str_split(lw, "CHAPTER ")[[1]][-1]

## 47 chapters
length(little_women) 

## Make the data frame requested in the question
chapters <- tibble(number=seq_along(chapter_names), title= chapter_names, text = little_women)
```



> The chapter number and name are in all caps at the begining of each chapter - remove the number and title from the `text` variable.

```{r chapter-names}
## Quick peak at beginning of text
ii <- 1
str_sub(chapters$text[[ii]],1,60)
str_sub(chapters$title[[ii]],1,60)


chapters <- chapters %>%
  mutate(title_length = str_length(title)) %>% #Add variable with length of titles
  mutate(text = str_sub(text, title_length+1, -1)) %>% #Remove title (plus white space)
  select(-title_length)

```




> Get the cumulative number of times the main characters' names appear by chapter

The main characters in Little Women are the sisters Amy, Beth, Jo, and Meg, and the neighbor boy Laurie.

```{r chapters-plot}
chapters <- chapters %>%
  mutate(amy = cumsum(str_count(text,"Amy"))) %>%
  mutate(beth = cumsum(str_count(text,"Beth"))) %>%
  mutate(jo = cumsum(str_count(text,"Jo"))) %>%
  mutate(laurie = cumsum(str_count(text,"Laurie"))) %>%
  mutate(meg = cumsum(str_count(text,"Meg"))) 
```


Next we will create a plot that displays the cumulative number of times the main characters are mentioned by chapter.  
In order to more easily color lines according to the characters, we will put the `chapters` data frame in long form.

```{r}
chapters_long <- chapters %>%
  pivot_longer(cols=amy:meg, names_to="character",values_to="count") 
  
ggplot(chapters_long, aes(x=number,y=count,color=character)) +
  geom_line(size=2)+
  xlab("Chapter number") +
  ylab("Cumulative count")+
  labs(color="Character")

```


Can this graphic help us to understand anything about the plot?

It turns out, yes!

A few observations:

- Jo is the principal character and her name appears regularly throughout the chapters

- Meg gets married

- Jo moves to NYC for a period and so the other characters are not mentioned as much

- Eventually Amy and Laurie get married 











