---
title: "Airport Waiting Times"
author: "Jessi Kehe"
output: html_document
---
This R Markdown document was adapted from documents created by Professor Bret Larget.

### Setup details

* You will need the packages `tidyverse`, `lubridate`, and `XML` for these lectures.  Make sure to install them if you haven't already.  

* This assumes you have the R script `viridis.R` two steps back from the working directory (`"../../scripts/viridis.R"`).  Be sure to adjust the code if you have this script in a different location.

* The following data files will be used and are assumed to be located two steps back from the working directory in a folder called `data/`.  Be sure to adjust the code if you have the data files in a different location.  
`"C:/stat_240/data/DEN-2018.html"`  
`"C:/stat_240/data/JFK-2018.html"`  
`"C:/stat_240/data/LAX-2018.html"`  
`"C:/stat_240/data/MSP-2018.html"`  
`"C:/stat_240/data/ORD-2018.html"`  
`"C:/stat_240/data/SFO-2018.html"`  

```{r setup, include=FALSE}
install.packages("XML")
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
library(lubridate)
library(XML)
## Note: this code assumes viridis.R is two steps up the file directory tree in a `scripts` folder
source("C:/stat_240/scripts/viridis.R")
```


# Lecture:  Importing the data files

## Airport Waiting Times

### Overview
* In the US, Customs and Border Patrol gathers statistics on the number of flights and passengers that go through customs at many US airports on an hourly basis.

* They also gather statistics on waiting times.
  - mean and maximum time  
  - number of booths open  
  - separate for domestic and international travelers  
  
###  Airports
* We will look at data from 2018 on several US Airports
  - DEN (Denver) 
  - JFK (New York)  
  - LAX (Los Angeles)  
  - MSP (Minneapolis)
  - ORD (Chicago, O'Hare)  
  - SFO (San Francisco)  
  
### R commands
* XML package  
  - `readHTMLTable()`
  
* dplyr package:  
  - `bind_rows()`
  - `mutate_f()`
  - `mutate_at()`
  
* lubridate package:  
  - `mdy()`
  

## Reading in an HTML Table

- The function `readHTMLTable()` in the `XML` package can read in tables from HTML files.
- However, it expects tables to have a single header row.
- When they do not, such as in this example, it takes more care.
- The function code below does a number of things to read in the data.
    - Grab part of a list as the data
    - Use `slice()` to get rid of the three header lines
    - Add new column names
    - Use `mutate_if()` and `mutate_at()` to change format of some variables
    - Use `mutate()` from `dplyr` and `mdy()` from `lubridate` to modify the date variable.
    
    
### Consider MSP for our test airport file 
```{r read-airport-html-1}
### Let's look at the actual HTML file first, then try to read in the data
test_airport_file <- "C:/stat_240/data/MSP-2018.html" # specify airport file
```


#### Specify the column names that we want to use
```{r read-airport-html-2}
# Next we specify the column names
airport_column_names = c("airport",
                           "terminal",
                           "date",
                           "hour",
                           "us_avg_wait",
                           "us_max_wait",
                           "non_us_avg_wait",
                           "non_us_max_wait",
                           "all_avg_wait",
                           "all_max_wait",
                           "all_n_0_15",
                           "all_n_16_30",
                           "all_n_31_45",
                           "all_n_46_60",
                           "all_n_61_90",
                           "all_n_91_120",
                           "all_n_120_plus",
                           "all_excluded",
                           "all_total",
                           "all_flights",
                           "all_booths")
airport_column_names
```

#### Read in the data using readHTMLTable()
```{r read-airport-html-3}
test_airport <- readHTMLTable(test_airport_file)$`NULL` %>% 
    slice(-(1:3)) # Remove the first three rows 

names(test_airport) # default column names
names(test_airport) <- airport_column_names # replace with new names
names(test_airport)
```
#### Make a Tibble and adjust structure

First look at the current structure - the variables are all set as characters.
```{r read-airport-html-4}
# Look at the data types for the different variables
str(test_airport)
```

Adjust the data types as needed.
```{r read-airport-html-5}
test_airport <- test_airport %>%
    as_tibble() %>%
    mutate_at(.vars = vars(ends_with("wait"),starts_with("all")),
              .funs = as.numeric) %>% # Turn the wait times into numbers
    mutate_if(is.factor,as.character) %>% # If any variables are factors, convert to characters
    mutate(date = mdy(date)) # Set `date` as a Date format; specify form of data (month, day, year)
str(test_airport)
```


### Define function for reading in other files

We will do through the same steps as with our test airport for several different data files.  
To make the code cleaner, let's define a function `read_airport()` that takes an HTML and carries out the same steps as above.
```{r stuff-for-all-airports}
read_airport <- function(x)
  # Input:  x = HTML file
  # Output: nicely formatted tibble
{
  airport_column_names <- c("airport",
                           "terminal",
                           "date",
                           "hour",
                           "us_avg_wait",
                           "us_max_wait",
                           "non_us_avg_wait",
                           "non_us_max_wait",
                           "all_avg_wait",
                           "all_max_wait",
                           "all_n_0_15",
                           "all_n_16_30",
                           "all_n_31_45",
                           "all_n_46_60",
                           "all_n_61_90",
                           "all_n_91_120",
                           "all_n_120_plus",
                           "all_excluded",
                           "all_total",
                           "all_flights",
                           "all_booths")

  airport <- readHTMLTable(x)$`NULL` %>%
    slice(-(1:3))

  names(airport) <- airport_column_names

  airport <- airport %>%
    as_tibble() %>%
    mutate_at(.vars = vars(ends_with("wait"),starts_with("all")),
              .funs = as.numeric) %>%
    mutate_if(is.factor,as.character) %>%
    mutate(date = mdy(date))
  
  return(airport)
}
```

## Read the Data

```{r read-data}
den <- read_airport("C:/stat_240/data/DEN-2018.html")
jfk <- read_airport("C:/stat_240/data/JFK-2018.html")
lax <- read_airport("C:/stat_240/data/LAX-2018.html")
msp <- read_airport("C:/stat_240/data/MSP-2018.html")
ord <- read_airport("C:/stat_240/data/ORD-2018.html")
sfo <- read_airport("C:/stat_240/data/SFO-2018.html")
```

## Combine the data into one large data frame

```{r combine-data}
awt <- bind_rows(den,jfk,lax,msp, ord,sfo) #airport waiting time
rm(den,jfk,lax,msp,ord,sfo)
View(awt)
str(awt)
```






# Lecture:  Exploratory Data Analysis

### Any missing data?

```{r missing-data}
count_na <- function(x)
{
  return ( sum(is.na(x) ) )
}

awt %>%
  summarize_all(count_na)

# There are no missing values for any variables.
```

### Summarize Key Variables

```{r key-variables-airport-passengers-flights}
## airports and terminals
awt %>%
  group_by(airport,terminal) %>%
  summarize(passengers = sum(all_total),
            flights = sum(all_flights))

### JFK and LAX have five different international terminals
### The other airports have one or two each.
```

```{r key-variables-airport-dates}
## examine date by airport
awt %>%
  select(airport,date) %>%
  distinct() %>%
  group_by(airport) %>%
  summarize(n = n())
## One date is missing for all the airports
```

```{r key-variables-missing-date}
## Let's figure out which date is missing
awt <- awt %>%
  mutate(month = month(date, label=TRUE)) %>% # add a month variable
  select(airport,terminal,date,month,everything())

awt %>%
  select(airport,date,month) %>%
  distinct() %>%
  group_by(airport,month) %>%
  summarize(n = n()) %>%
  filter(month == "May")
  

```


```{r key-variables-may}
awt %>%
  select(airport,date,month) %>%
  filter(month == "May") %>%
  distinct() %>%
  group_by(date) %>%
  summarize(n=n())

## check on May 8
awt %>%
  filter(date == "2018-05-08") %>%
  summarize(n=n())

### No airports have any data on May 8, 2018.
```

#### Check totals

```{r key-variables-all_total}
## check that all_total column matches
awt <- awt %>%
  mutate(check_total = all_n_0_15 +
            all_n_16_30 +
            all_n_31_45 +
            all_n_46_60 +
            all_n_61_90 +
            all_n_91_120 +
            all_n_120_plus +
            all_excluded)

awt %>%
  summarize(diff = sum(all_total != check_total))

table_mismatch <- awt %>%
  filter(all_total != check_total)
table_mismatch
```


```{r key-variables-all_total-times}
table_mismatch %>%
  group_by(hour) %>%
  summarize(n = n())
### All mismatches are between midnight and 1am.
```

```{r key-variables-all_total-mismatch-graphic}
awt %>%
  filter(all_total != check_total) %>%
ggplot(aes(x=all_total - check_total)) +
  geom_histogram(boundary=0,binwidth=5)
```

#### Let's consider another way to check the totals (`all_total`) using `purrr::pmap_dbl()`.

```{r add-variables}
awt <- awt %>%
  ## add day of the week
  mutate(wday = wday(date, label=TRUE)) %>%
  ## check total obvious way 
  mutate(check_total = all_n_0_15 +
            all_n_16_30 +
            all_n_31_45 +
            all_n_46_60 +
            all_n_61_90 +
            all_n_91_120 +
            all_n_120_plus +
            all_excluded) %>%
  ## check total in a new way using purrr::pmap_dbl()
  ## pmap_dbl here takes a data frame as its first argument
  ##   and then applies the function row-wise to its elements, we will use the `sum()` function
  ## select() supplies the data frame 
  ## because we do not pipe into select here directly (we're inside of a pipe into mutate),
  ##   we need to specify the input to select (we need to note which variables we want to sum over)
  ##   The '.' represents the object that was piped into mutate
  ##   To this we select the columns that contain "_n_" and 'all_excluded'
  ## the row sums of this selected sub-data frame are saved as a new
  ##   variable named check_total2
  mutate(check_total2 =
           pmap_dbl(select(.,contains("_n_"),all_excluded),sum)) %>%
  select(airport,terminal,date,month,wday,everything())

## Verify that the check sums match each other
awt %>%
  select(check_total,check_total2) %>% # two variables totaling the passenger counts
  mutate(diff = check_total - check_total2) %>% # take the difference
  summarize(min = min(diff),
            max = max(diff),
            ndiff = sum(diff != 0))
## Note that there are no differences

## Compare check_total to all_total
check_total_mismatch <- awt %>%
  filter(check_total != all_total) %>%
  mutate(error = all_total - check_total) %>%
  select(airport,date,hour,all_total,check_total,error,everything())

print(paste0(nrow(check_total_mismatch), " observations out of ", nrow(awt), " have an error, or ", round(nrow(check_total_mismatch)/nrow(awt)*100, 2), " %"))

##---will use check_total instead of all_total, but likely will not matter.
```

#### Summary of findings

The data do not contain any observations on May 8, 2018.

The errors in `all_total` all involve a few counts with totals larger than the checked total (`check_total`). We can use `check_total` as the actual value, but this assumes the individual column counts are correct. Some correction or ignoring the issue is not likely to make much difference in major questions.

### Associations

```{r associations}
ggplot(awt, aes(x=us_avg_wait,y=us_max_wait,color=airport)) +
  geom_point() +
  geom_abline(intercept=0,slope=1) +
  facet_wrap(~airport)

ggplot(awt, aes(x=non_us_avg_wait,y=non_us_max_wait,color=airport)) +
  geom_point(alpha=.05) +
  geom_abline(intercept=0,slope=1) +
  facet_wrap(~airport)

ggplot(awt, aes(x=us_avg_wait, y=non_us_avg_wait, color=airport)) +
  geom_point(alpha = 0.5) +
  geom_abline(intercept=0,slope=1) +
  facet_wrap(~airport)

## It looks like there are some times where there were few or no non-US international travelers
awt %>%
  filter(non_us_avg_wait == 0)
```


## Some questions to consider

1. Are some airports more efficient that than others in terms of average wait time?  
    - Are some airports more efficient in the number of people served by hour per available booth?

2. Are some months busier than others?  
    - Are some days of the week busier than others?  
    - Are some times during the day busier than others?  

3. How does the proportion of people who get through customs in less than 30 minutes depend on other factors?  


# Lecture:  Airport efficiency


> Are some airports more efficient than others in terms of average wait time?

```{r efficient}
ggplot(awt, aes(x=airport, y=all_avg_wait, color=airport)) +
  geom_boxplot()

awt %>%
  mutate(airport = reorder(airport,-all_avg_wait, FUN=median)) %>% 
  #`airport` as factor with levels ordered according to -all_avg_wait and FUN value
  ggplot(aes(x=airport, y=all_avg_wait, color=airport)) +
  geom_boxplot() +
  ggtitle("Are some airports more efficient that than others in terms of average wait time?")
```


> Are some airports more efficient in the number of people served by hour per available booth?

```{r people-per-hour-per-booth}
awt %>%
  group_by(airport, hour) %>%
  summarize(total = sum(check_total), booth_total=sum(all_booths)) %>%
  mutate(per_booth=round(total/booth_total,2)) %>%
  select(-total, -booth_total) %>%
  ggplot(aes(hour, per_booth, color=airport)) +
  geom_point() +
  geom_hline(yintercept=45, linetype="dashed",color="red") +
  facet_wrap(~airport) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Passengers per booth") +
  ggtitle("Are some airports more efficient", 
          subtitle="in the number of people served by hour per available booth?")
```

There do appear to be differences between the efficiency (in terms of passengers per booth per hour) among the airports.  For example, MSP seems to be more efficient (in the noted sense) regularly serving more than 45 passengers per booth per hour. SFO seems to serve fewer than 45 passengers per booth per hour most of the time.







# Lecture:  Busy times

> Are some months busier than others?

Count the total number of passengers by month and sort.
```{r months}
awt %>%
  group_by(month) %>%
  summarize(total = sum(check_total)) %>%
  arrange(desc(total))
```

The previous summary is affected by different numbers of days in each month.
Calculate the average daily rate of passengers by month.
(Recall that May has only 30 observations.)

Key idea here is to first summarize by month and day, and then by month,
in order to get the total number of passengers in each month
and the number of days in each month.

```{r busy-months}
busy_months <- awt %>%
  mutate(day = day(date)) %>%
  group_by(month,day) %>%
  summarize(total = sum(check_total)) %>%
  group_by(month) %>%
  summarize(n=n(), total = sum(total), daily = total/n) %>%
  arrange(desc(daily))
busy_months

ggplot(busy_months, aes(month, daily)) +
  geom_point() +
  xlab("Day of the week") +
  ylab("Average number of passengers per day") + 
  ggtitle("Are some months busier than others?")
```

It looks like the summer months (July, August, June) are the busiest in terms of number of passengers traveling per day for the airports in our data.


> Are some days of the week busier than others?  

```{r busy-weekdays-incorrect}
# This approach is wrong - can you see why?
# Look at the displayed `total` for each day...were there 6,544,008 Saturdays in 2018?!
# This is count is based on the rows in the table with wday == "Sat", not the number
# of Saturdays in 2018.
awt %>%
  group_by(wday) %>%
  summarize(total = sum(check_total), n = n()) %>%
  mutate(daily=round(total/n,2)) %>%
  arrange(desc(daily))
```


```{r busy-weekdays}
# We first need to count the number of each day of the week in 2018
busy_wdays <- awt %>%
  group_by(date, wday) %>%
  summarize(total = sum(check_total)) %>%
  group_by(wday) %>%
  summarize(n=n(), total = sum(total))  %>%
  mutate(daily=round(total/n,2)) %>%
  arrange(desc(daily)) 
busy_wdays

ggplot(busy_wdays, aes(wday, daily)) +
  geom_point() +
  xlab("Day of the week") +
  ylab("Average number of passengers") +
  ggtitle("Are some days of the week busier than others?")
```





> Are some times during the day busier than others?  

```{r busy-times}
busy_times <- awt %>%
  group_by(date, hour) %>%
  summarize(total = sum(check_total)) %>%
  group_by(hour) %>%
  summarize(n=n(), total = sum(total))  %>%
  mutate(hourly=round(total/n,2)) %>%
  arrange(desc(hourly)) 
busy_times

ggplot(busy_times, aes(hour, hourly)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))  +
  xlab("Time of the day") +
  ylab("Average number of passengers") +
  ggtitle("Are some times during the day busier than others?")
```







# Lecture:  Getting through customs quickly

> How does the proportion of people who get through customs in less than 30 minutes depend on other factors?  

```{r add-proportions}
## Add columns for proportion less than 30 minutes and proportion more than an hour
awt <- awt %>%
  mutate(p_less_30 = (all_n_0_15 + all_n_16_30) / check_total,
         p_more_60 = (all_n_61_90 + all_n_91_120 + all_n_120_plus) / check_total)
```


#### Factor: Airport

```{r less30-airport}
awt %>%
  mutate(airport = reorder(airport,p_less_30,FUN=median)) %>%
ggplot(aes(x=airport,y=p_less_30,color=airport)) +
  geom_boxplot() +
  ggtitle("Proportion of people through customs in < 30 minutes depend on other factors?",
          subtitle = "Factor: airport")
```

#### Factor: Hour

```{r less30-hour}
## Fraction of passengers per hour that go through in < 30 minutes
awt %>%
  group_by(hour) %>%
  summarize(total = sum(check_total),
          n_less_30 = sum(all_n_0_15) + sum(all_n_16_30),
          p_less_30 = n_less_30 / total,
          booths = sum(all_booths)) %>%
  ggplot(aes(hour,p_less_30, size = booths)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Proportion of people through customs in < 30 minutes depend on other factors?",
          subtitle = "Factor: hour")
```



#### Factor: Airport and Hour

```{r less30-airport-hour}
## Fraction of passengers per airport and per hour that go through in < 30 minutes
awt %>%
  group_by(airport,hour) %>% 
  summarize(total = sum(check_total),
            n_less_30 = sum(all_n_0_15) + sum(all_n_16_30),
            p_less_30 = n_less_30 / total,
            booths = sum(all_booths)) %>%
  ungroup() %>% ## airport is a grouping variable and cannot be mutated
  mutate(airport = reorder(airport,p_less_30,FUN=mean)) %>%
  ggplot(aes(x=airport, y=p_less_30,fill=airport)) +
  geom_col() +
  facet_wrap(~hour) + 
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  ggtitle("Proportion of people through customs in < 30 minutes depend on other factors?",
          subtitle = "Factor: airport and hour")
```

```{r less30-airport-hour-check}
# It looks like some of the airports do not have passengers in certain time windows
# Is that correct?  
awt %>% 
  filter(airport=="MSP") %>%
  group_by(hour) %>%
  summarize(tot = sum(check_total))
```


```{r less30-airport-hour-2}
## Let's try a scatterplot of the same information as the previous graphic
## -> which is easier to interpret?
## Fraction of passengers per airport and per hour that go through in < 30 minutes
awt %>%
  group_by(airport,hour) %>% 
  summarize(total = sum(check_total),
            n_less_30 = sum(all_n_0_15) + sum(all_n_16_30),
            p_less_30 = n_less_30 / total,
            booths = sum(all_booths)) %>%
  ggplot(aes(x=hour, y=p_less_30,color=airport, shape=airport)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Proportion of people through customs in < 30 minutes depend on other factors?",
          subtitle = "Factor: airport and hour")
```







#### Factor: Number of Booths and Hour

```{r less30-booth-hour}
# Look at passengers per booth vs. hour
# Color based on proportion who wait less than 30 minutes
# Consider size based on the number of booths open
awt %>%
  group_by(airport,hour) %>% 
  summarize(total = sum(check_total),
            n_less_30 = sum(all_n_0_15) + sum(all_n_16_30),
            p_less_30 = n_less_30 / total,
            booths = sum(all_booths)) %>%
  mutate(passenger_per_booth = total/booths) %>%
  ggplot(aes(hour,passenger_per_booth, color=p_less_30, size = booths)) +
  geom_point() +
  geom_hline(yintercept=45, color="red", linetype="dashed") +
  facet_wrap(~airport) +
  ylab("Number of passengers per available booth") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Proportion of people through customs in < 30 minutes depend on other factors?",
          subtitle = "Factor: booth and hour")
```



```{r less30-booth-hour-2}
# Look at proportion who wait less than 30 minutes vs. hour
# Color based on proportion number of passengers per booth
awt %>%
  group_by(airport,hour) %>% 
  summarize(total = sum(check_total),
            n_less_30 = sum(all_n_0_15) + sum(all_n_16_30),
            p_less_30 = n_less_30 / total,
            booths = sum(all_booths)) %>%
  mutate(passenger_per_booth = total/booths) %>%
  ggplot(aes(hour,p_less_30, color=passenger_per_booth)) +
  geom_point() +
  facet_wrap(~airport) +
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ggtitle("Proportion of people through customs in < 30 minutes depend on other factors?",
          subtitle = "Factor: booth and hour")
```



> Can you think of your own questions to address with this data set?




