---
title: "readr and lubridate"
author: "Jessi Kehe"
output: html_document
---

### Setup details

* You will need the packages **tidyverse** and **lubridate** for this lecture. 

* Put data file at "../../data/readr_example.txt"

* Be sure to read R4DS Chapter 11 for information on **readr** and Chapter 16 for information on **lubridate**

* We are going to consider a few simple examples to help you get started with **readr** and **lubridate**.  Practice on more complicated examples will come as the semester progresses.
  
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,message=FALSE,warning=FALSE,cache=TRUE,autodep=TRUE,cache.comments=FALSE)
library(tidyverse)
library(lubridate)
```


# readr - importing data

Depending on the format of a data file, different functions are needed to read the data file into and produce a data frame.  Some common options are listed below.

`read_csv()` for comma delimited files 

`read_csv2()` for semicolon separated files

`read_tsv()` for tab delimited files

`read_delim()` for in files with any delimiter


```{r readr}
df1 <- read_delim("C:/stat_240/week_4/readr_example.txt", delim="$")
df1
df2 <- read_delim("C:/stat_240/week_4/readr_example.txt", delim="!")
df2
df3 <- read_csv("C:/stat_240/week_4/readr_example.txt")
df3

colnames(df1)
colnames(df1) <- c("a", "b", "c", "d")
colnames(df1)

df4 <- read_delim("C:/stat_240/week_4/readr_example.txt", delim="$", skip=1) #let's skip one row
df4

df5 <- read_delim("C:/stat_240/week_4/readr_example.txt", delim="$", skip=1, col_names = FALSE)
df5
colnames(df5) <- c("a", "b", "c", "d")
df5
```


# lubridate - working with dates and times

#### Dates from strings

```{r date-from-string}
# Uses the ISO 8601 format where components of a specified date are arranged 
# from biggest to smallest: year, month, day, hour, minute, second
date1 <- "February 5 2017"
class(date1)
mdy(date1)
class(mdy(date1))
```

#### More examples

```{r date-from-string-examples}
date2 <- "5-2-2017"
date3 <- "2017, February 5"
date4 <- "2.2017.5"

dmy(date2)
mdy(date2) # changed month and day
ymd(date3)
myd(date4)
```

#### Components of dates

```{r date-components}
test_date <- mdy(date1)
test_date

year(test_date)
month(test_date)
month(test_date, label=TRUE)
month(test_date, label=TRUE, abbr=FALSE)
day(test_date)
wday(test_date)
wday(test_date, label=TRUE)
wday(test_date, label=TRUE, abbr=FALSE)
```







