---
title: "Madison Weather - Overview"
author: "Jessi Kehe"
output: html_document
---

This R Markdown document was adapted from documents created by Professor Bret Lager.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
install.packages("viridis")
library(tidyverse)
library(lubridate)

## reset color defaults
## Source https://data-se.netlify.com/2018/12/12/changing-the-default-color-scheme-in-ggplot2/
library(tidyverse)
library(viridis) # new colors
library(scales)

#### continuous variables color and fill

options(ggplot2.continuous.colour="viridis")
options(ggplot2.continuous.fill = "viridis")

#### use viridis for discrete scales
scale_colour_discrete <- scale_colour_viridis_d
scale_fill_discrete <- function(...) {
   scale_fill_manual(..., values = viridis_qualitative_pal7)
} 

## reset default theme
theme_set(theme_minimal()) # ?:A minimalistic theme with no background annotations.
```

## Overview of Madison Weather Data

> The source of the data for this week is from the National Oceanic and Atmospheric Administration (NOAA) of the United States.

The data file is `madison-weather-2020.csv`, and the variable descriptions are provided below.

STATION: unique code for the weather station   
NAME: station name    
LATITUDE: station latitude  
LONGITUDE: station longitude  
ELEVATION: station elevation (feet above sea level)   
DATE: date of observations  
AWND: average wind speed (mph)  
PRCP: precipitation (inches, water)  
SNOW: snow fall (inches)  
SNWD: snow depth (inches)  
TAVG: daily average air temperature (degrees Fahrenheit)   
TMAX: maximum air temperature (degrees Fahrenheit)   
TMIN: minimum air temperature (degrees Fahrenheit)   
WSF1: maximum one-minute wind speed  

### Import and Transform Data

The data file has one row for each date and weather station.
The following block of code will read in the data,
eliminate two of the weather stations,
eliminate many of the variables,
and then combine the data from the remaining data sets so that there is just one observation per date.

```{r import-transform-data}
## Read in the data
## Specify the type
## This is needed to avoid issues for columns where first 1000 rows are all missing
##   and the type is set to logical
mw <- read_csv("madison-weather-2020.csv",
              col_types = cols(
  STATION = col_character(),
  NAME = col_character(),
  LATITUDE = col_double(),
  LONGITUDE = col_double(),
  ELEVATION = col_double(),
  DATE = col_date(format = ""),
  AWND = col_double(),
  PRCP = col_double(),
  SNOW = col_double(),
  SNWD = col_double(),
  TAVG = col_double(),
  TMAX = col_double(),
  TMIN = col_double(),
  WSF1 = col_double()
))


## recode the NAME variable
mw <- mw %>%
  mutate(NAME = recode(NAME,
                       `ARBORETUM UNIVERSITY WIS, WI US` = "Arboretum",
                       `CHARMANY FARM, WI US` = "Charmany",
                       `MADISON DANE CO REGIONAL AIRPORT, WI US` = "Airport",
                       `MADISON MUNICIPAL AIRPORT, WI US` = "Municipal Airport",
                       `MADISON WEATHER BUREAU CITY, WI US` = "Bureau"))

## Select a few columns and reduce to two weather stations
## Rename the variables with lower case letters
mw <- mw %>%
  select(NAME,DATE,PRCP,TMAX,TMIN) %>%
  filter(NAME == "Airport" | NAME == "Bureau") %>%
  rename(name = NAME,
         date = DATE,
         prcp = PRCP,
         tmax = TMAX,
         tmin = TMIN)

## Find earliest and latest date
mw %>%
  group_by(name) %>%
  summarize(min = min(date),max=max(date))

## Decide to use all Airport data
##  and only use Bureau data through 1939-09-30
mw <- mw %>%
  filter(name == "Airport" | date < "1939-10-01")

## Double check
mw %>%
  group_by(name) %>%
  summarize(min = min(date),max=max(date))

## Check missing data
mw %>%
  summarize_all(function(x) sum(is.na(x)))

## Drop rows with missing data
mw <- mw %>%
  drop_na()

## Calculate average temperature and precipitation by year
## Drop partial year of 2020
## Add a period for 30-year periods from 1870 through 2019
breaks = seq(1869,2019,30)
labels = str_c((breaks+1)[-6],breaks[-1],sep="-") #?str_c: Joins two or more vectors element-wise into a single character vector, optionally inserting sep between input vectors.
mw_year = mw %>%
  mutate(tavg = (tmin+tmax)/2) %>%
  mutate(year = year(date)) %>%
  filter(year < 2020 & year > 1869) %>%
  group_by(year) %>%
  summarize(n = n(),
            tmin = mean(tmin),
            tavg = mean(tavg),
            tmax = mean(tmax),
            prcp = sum(prcp)) %>%
  mutate(period = cut(year,
                      breaks = breaks,
                      labels = labels))
## Summarize by month and period
mw_month = mw %>%
  mutate(tavg = (tmin+tmax)/2) %>%
  mutate(month = month(date, label=TRUE)) %>%
  mutate(year = year(date)) %>%
  filter(year < 2020 & year > 1869) %>%
  mutate(period = cut(year,
                      breaks = breaks,
                      labels = labels)) %>%
  group_by(period,month) %>%
  summarize(n=n(),
            tmin = mean(tmin),
            tmax = mean(tmax),
            tavg = mean(tavg),
            prcp = sum(prcp))
```

### Graph of minimum temp over time

```{r tmin}
ggplot(mw_year, aes(x=year, y=tmin)) +
  geom_point() +
  geom_smooth() +
  xlab("Year") +
  ylab("Average Minimum Temperature (degrees F)") +
  ggtitle("Madison Annual Temperature",
          subtitle = "1870-2019")
```

```{r tmax}
ggplot(mw_year, aes(x=year, y=tmax)) +
  geom_point() +
  geom_smooth() +
  xlab("Year") +
  ylab("Average Maximum Temperature (degrees F)") +
  ggtitle("Madison Annual Temperature",
          subtitle = "1870-2019")
```

```{r tavg}
ggplot(mw_year, aes(x=year, y=tavg)) +
  geom_point() +
  geom_smooth() +
  xlab("Year") +
  ylab("Average Annual Temperature (degrees F)") +
  ggtitle("Madison Annual Temperature",
          subtitle = "1870-2019")
```

```{r tmax-boxplot}
ggplot(mw_year, aes(x=period, y=tmax, color=period)) +
  geom_boxplot() +
  geom_point(position = position_jitter(width=0.1,height=0)) +
  xlab("30-Year Period") +
  ylab("Average Maximum Temperature (degrees F)") +
  ggtitle("Madison Annual Temperature by Period",
          subtitle = "1870-2019")
```

#### Precipitation Plots

```{r precip}
ggplot(mw_year, aes(x=year, y=prcp)) +
  geom_point() +
  geom_smooth() +
  xlab("Year") +
  ylab("Annual Total Precipitation (inches)") +
  ggtitle("Madison Annual Precipitation",
          subtitle = "1870-2019")
```

#### Period Month Plot

```{r period-month}
ggplot(mw_month, aes(x=month,y=tavg,color=period)) +
  geom_point() +
  facet_wrap(~period) +
  theme(axis.text.x = element_text(angle=45))
```








