---
title: "Discussion 1"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
library(tidyverse)
```

## Preliminaries

Read the instructions.  
Create discussion folders and change browser settings if needed.  
Download the files `01-discussion-basics.Rmd` and `lake-mendota-clean-2020.csv` to your `COURSE/Discussion/01-Discussion/` directory.

## Questions

### Question 1

Set the `01-Discussion` directory to be be your *working directory*.
Type `getwd()` in the Console and verify that you see that the working directory is set to be `01-Discussion`.
If not, learn how to set the working directory from someone in your group or the TA.

When you knit this document,
the following R chunk will print your working directory to the output document.

```{r getwd, echo=TRUE, results='asis'}
getwd()
```

### Question 2

The following R chunk will read the data.
The `names()` command will print the names of the variables in the data frame.

Replace the text XXX below with the correct numbers.

```{r read-data}
mendota <- read_csv("lake-mendota-clean-2020.csv")
names(mendota)
```

#### Response

The Lake Mendota data set has XXX variables and the first variable (next to the `[1]`) is named XXX.


## Submission

Once you have completed all of the questions, knit the document.
This should create an HTML document in your `01-Discussion` folder.

Go to the Canvas course web page, find the Assignments section in the left panel, and upload both your edited `01-discussion-basics.Rmd` and `01-discussion-basics.html` files.