---
title: "Lake Mendota Freezing and Thawing"
author: "Jessi Cisewski Kehe"
date: "Fall 2020:  Week 1"
output: html_document
---

This R Markdown document was adapted from a document created by Professor Bret Lager.

```{r setup, include=FALSE}
## <- These hash tags allow you to comment-out the code; you can use multiple depending on your preference but only one is needed.
## 'knitr' is the package that turns R Markdown into a document
## The following command sets some options for this R Markdown file
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE)
## The `library()' function loads in a package; the package must already be installed.
## `tidyverse' is the name of the package we are loading next.  In particular, we will be using ggplot2.
library(tidyverse)

## We will also load the package called `lubridate.'  It is a part of `tidyverse', but it is not loaded in the core.  
##  (This means we need to load it separately when we want to use it in our code.)
## The 'lubridate' package has useful functions for dates.
library(lubridate)
```


### Read the data

The following R chunk has one line of code that will read in the data set.
Along with the code I added many comments to explain what is happening.

Each code chunk begins with three consecutive back ticks at the start of a line.

The left brace `{` followed by the letter `r` means that the code should be interpreted as r code.

The name `read-data` is the name of the chunk.
Naming chunks is helpful when trying to find errors.
Other knitr options could be set before the right brace `}`, but are not in this example.

```{r read-data}
## `ice' is a new object. 
## The `<-' assigns the loaded data set to the object name `ice.'
## You will see an '=' instead of a `<-' as the assignment operator; I use both, but many R books recommend the `<-.'
## read_csv() is a function in the tidyverse for reading in .csv files.
## There is a base R function named read.csv(). Use read_csv() instead.
## The argument to read_csv() is a path to the file with the data
## The '..' means go up a directory
## Use a '/' after a directory
##   (Windows uses backslashes, but slashes work in R for any computer system)
## The result of the code below is to create a data frame named ice
##   by reading in the data in the file.
ice <- read_csv("C:/stat_240/lake-mendota-year-clean.csv")
```

##### str( ) and spec( )

The R command `str()` is useful to check the structure of a data set (or other R objects).
This code displays the structure of the data set.

The command `spec()` is part of the tidyverse and displays how each variable is specified after reading in.

```{r}
str(ice)
spec(ice)
```

### Plots

#### One-variable Plots

The variable *days* is the number of days that Lake Mendota was frozen each winter.

This variable can be displayed with many different types of graphs.
Every plot we make using the **ggplot2** package starts with the function `ggplot()` which takes the name of a data frame as its first argument.
Then, new layers can be added to the plot.
This almost always includes at least one geometric element, a function that begins with `geom_` followed by a descriptive name.
Most plots will include multiple layers to add labels, modify previous layers, add more features to the plot, and so on.
Note that the `+` sign must be at the end of each line, not the beginning of the line.
(If not, R thinks you were done adding layers the first time there is not a plus-sign at the end.)

#### Histogram

We can use ggplot to make a basic histogram like the following.

```{r histogram}
ggplot(ice) +
  geom_histogram(aes(x=days))
```

This histogram changes 

  * the labels with  `ggtitle()`, `xlab()`, and  `ylab()`
  
  * the widths of the bars with the `binwidth=` option 
  
  * the boundaries where the bars are set with the `boundary=` option 
  
  * the color of the bar interiors with the `fill=` option
  
  * the color of the bar outlines with the `color=` option

```{r histogram-fancy}
ggplot(ice) +
  geom_histogram(aes(x=days),
                 fill="hotpink",
                 color="black",
                 binwidth=7,
                 boundary=0) +
  xlab("Total days frozen") +
  ylab("Counts") +
  ggtitle("Lake Mendota Freeze Durations, 1852-2019")


ggplot(ice) +
  geom_histogram(aes(x=days),
                 fill = "blue",
                 color = "black",
                 binwidth = 9,
                 boundary=0) +
  xlab("Total days frozen") +
  ylab("Counts") +
  ggtitle("Lake Mendota Freeze Durations, 1852-2019")
```

#### Density Plot

A density plot can be thought of as a smoothed histogram.  
We can display a density plot using `geom_density().`

```{r density}
ggplot(ice) +
  geom_density(aes(x=days),
               fill="hotpink",
               color="black") +
  xlab("Total days frozen") +
  ylab("Density") +
  ggtitle("Lake Mendota Freeze Durations, 1852-2019")
```

#### Boxplot

A boxplot requires a categorical variable for the x argument
and a quantitative variable for the y argument.

Each box outlines the middle 50% of the observations in the corresponding category:

* the bottom of the box (i.e., the lower hinge) gives the 25th percentile (the first quartile, Q1)

* the middle bar give the median (the 50th percentile)

* the top of the bax (i.e., the upper hinge) gives the 75th percentile (the third quartile, Q3)

The lower vertical line (i.e., the lower whisker) reaches down to the minimum value of the observations, but
will not drop below Q1 - 1.5 x IQR, where IQR = Q3 - Q1 is the inter-quartile range.

Similarly, the higher vertical line (i.e., the upper whisker) reaches up to the maximum value of the observations, but
will not go above Q3 + 1.5 x IQR.

Any points plotted below or above the vertical lines indicate observations that are 
below or above the noted ranges.  These points may be considered outliers.


```{r boxplot}
ggplot(ice) +
  geom_boxplot(aes(x=as.character(decade),y=days),
               fill="hotpink",
               color="black") +
  ylab("Total days frozen") +
  xlab("Decade") +
  ggtitle("Lake Mendota Freeze Durations, 1852-2019")
```

### Scatter Plot with a Trend Line Added

To create a scatter plot you can designate the `x` and `y` variables in `aes()`.
Then add points using geom_point().
If you want the points connected, you can also include `geom_line()`.
Using `geom_smooth()` will add a smooth line to the plot, which adapts to the data.
```{r plot}
ggplot(ice, aes(x=year1,y=days)) +
  geom_point() +
  geom_line() +
  geom_smooth(se=FALSE, span=.1) + #span =  #method="lm"
  xlab("Year") +
  ylab("Total Days Frozen") +
  ggtitle("Lake Mendota Freeze Durations, 1852-2019")
```