---
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE)
```

## Assignment 1

#### Due Friday, September 4, 11:59 PM

The purpose of this assignment is to verify that you have correctly installed R, R Studio, and the tidyverse.

### Preliminaries

Follow the instructions in the Course Notes to set up your computer.

[https://bookdown.org/bret_larget/stat-240-case-studies/](https://bookdown.org/bret_larget/stat-240-case-studies/)

#.  Install R
#.  Install RStudio
#.  Set preferences in RStudio
#.  Install the *tidyverse* package

### Problems

  1. Type the following commands into the console.

```{r, eval=FALSE}
set.seed(2020)
rpois(1, 25)
```

The output will be the characters `[1]` followed by a number. What is the number?

  2. Type the following commands into the console.
 
```{r, eval=FALSE}
library(tidyverse)
data(trees)
summarize(trees, n=n(), total=sum(Height))
```

The output will be a table that indicates the number of rows (n) of the data set `trees` and the sum of the variable `Height`. What is the sum of the variable `Height`, listed under the word **total** in the output?
  
### Submission

To submit this assignment, go to our Canvas site and select "Assignments" on the left panel.  Click the link for this assignment and follow the directions; there will be space for text entry of your responses. 
  
  
